---
title: "Chess Project Part 2"
author: "LonghaoChen&SenhWang"
date: "1/4/2022"
output: html_document
---

```{r setup, include=FALSE}
library(DBI)
library(dbplyr)
library(dplyr)
library(RSQLite)
library(tidyverse)
setwd("/Users/longhaochen/chess/chess/Database/")
db<-dbConnect(SQLite(),dbname="chess_project.db")
```


```{r}
dbListTables(db)

dbGetQuery(db, 'SELECT * FROM Match LIMIT 5')

match <- tbl(db, "Match")

match %>%
    select(everything())

head(match, 5)
```

1. What opening are most common? In different level of players?

```{r}
# We are interested in the most popular opening for each rank of players.
common_opening <- dbGetQuery(db, 'SELECT COUNT(*), opening, Rank FROM Match GROUP BY opening, Rank')

# First we sorted by most popular opening. We then wanted to know what the most popular opening is for each rank of players so that we can make the best opening recommendations for players to learn.
common_opening_2 <- match %>%
  select(ID, opening, Rank) %>%
  group_by(opening, Rank) %>%
  summarise(num_opening = n_distinct(ID) )  %>% 
  ungroup() %>%
  group_by (Rank) %>% 
  mutate(popularity = dense_rank(desc(num_opening)))%>% 
  arrange(desc(num_opening))

print(common_opening_2)

# Translates dplyr function to SQL query
show_query(match %>%
  select(ID, opening, Rank) %>%
  group_by(opening, Rank) %>%
  summarise(num_opening = n_distinct(ID) )  %>% 
  ungroup() %>%
  group_by (Rank) %>% 
  mutate(popularity = dense_rank(desc(num_opening)))%>% 
  arrange(desc(num_opening)))

#Visualize the result
result<-data.frame(common_opening_2) %>% filter(popularity <= 10)

l <- ggplot(result, aes(Rank, opening))
l + geom_tile(aes(fill = num_opening))
```
There are a few openings that are popular among all rank levels. For example, the Sicilian Defense and Scandinavian Defense: Mieses-Kotroc Variation openings are played by both beginners and experts. However, there also differences that we noticed. We can see that the Van\'t Kruijs Opening is popular among lower-ranked players and not as much among higher-ranked players. The Modern Defense and Indian Game openings are mostly played by experts and not by beginners.
Rank A: top 25% by Elo rating, Rank B: between 25-50% by Elo, Rank C: bottom 25-50% by Elo, Rank D: bottom 25% by Elo
num_opening is popularity of opening from greatest to least

2. What opening tends to end early (number of moves are low)?

```{r}
#First step is to sum up the steps in match_move. Then join the result with match. Last to group by different opening and level of players.



dbGetQuery(db, 'SELECT AVG(num_moves),COUNT(ID) as num_match,opening,rank FROM (SELECT FLOOR(COUNT(match_move.ID)/2) AS "num_moves", match_move.ID, match.rank, match.opening FROM Match_Move  join match on match_move.id = match.id GROUP BY match_move.id) GROUP BY opening, rank ORDER BY num_match desc')


```

3. What is the highest Elo in a game?
4. Give me a table so I can analyze the relationship between time usage (seconds) and eval? My null hypothesis is that if a player spend significant more time on a single move, the evaluation will improve!
5. What is the winning percentage of white? Also what is the percentage of games that end in draw? In different openings.